Towards Compiler-Agnostic Performance in Finite-Difference Codes. (TBC)

### 2014-11-19 docs/papers/shallow_performance/shallow_perf.tex
Add schematic of vanilla psykal shallow

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Do not use the stfloats baselinefloat ability as IEEE does not allow
% \baselineskip to stretch. Authors submitting work to the IEEE should note
% that IEEE rarely uses double column equations and that authors should try
% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
% packages (also by Sigitas Tolusis) as IEEE does not format its papers in
% such ways.
% Do not attempt to use stfloats with fixltx2e as they are incompatible.
% Instead, use Morten Hogholm'a dblfloatfix which combines the features
% of both fixltx2e and stfloats:
%
\usepackage{dblfloatfix}
% The latest version can be found at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/dblfloatfix/

@@

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/url/
% Basically, \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

@@

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{T}{his} demo file is intended to serve as a ``starter file''
for IEEE journal papers produced under \LaTeX\ using
IEEEtran.cls version 1.8 and later.

\subsection{The `shallow' Program}
For this work we have used a benchmarking program that solves the
shallow-water equations on a bi-periodic plane following the
finite-difference scheme introduced by Sadourney~\cite{sadourny75}.
Originally written by Paul Swarztrauber of the National Center for
Atmospheric Research, US.  Organic development - optimised for
previous hardware. In describing our work, we term the version of the
shallow code obtained at the beginning of this work the `original'
version.

@@

We do retain I/O because any realistic code has to output results to disk.

@@

@@

With the Gnu compiler (first and third clusters in
Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal version of
shallow only achieves ~5-6\% of the performance of the
original. Simply copying the code for each of the kernel subroutines
into the module from which they are called has the most dramatic
effect on the performance of the compiled code; it now achieves almost
80\% of the performance of the original. The next most significant
transformation is that of (manually) in-lining the kernel code itself into
the bodies of the loops from which they are called. This gets the
\psykal version to within 8\% of the performance of the original.
{\bf What's losing us 8\% of performance with the Gnu compiler?}

@@

! Copy new fields into current fields
DO J=1,N+1
  DO I=1,M+1
    Ufld(I,J) = UNEW(I,J)
  END DO
END DO
\end{verbatim}
\caption{Example of the coding structure that required manual
  in-lining of the kernel body (as indicated by commented-out lines)
  to retrieve performance of original shallow with the Intel
  compiler.}
\label{FIG_time_smooth_code}
\end{figure}

>>>>

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Do not use the stfloats baselinefloat ability as IEEE does not allow
% \baselineskip to stretch. Authors submitting work to the IEEE should note
% that IEEE rarely uses double column equations and that authors should try
% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
% packages (also by Sigitas Tolusis) as IEEE does not format its papers in
% such ways.
% Do not attempt to use stfloats with fixltx2e as they are incompatible.
% Instead, use Morten Hogholm'a dblfloatfix which combines the features
% of both fixltx2e and stfloats:
%
\usepackage{dblfloatfix}

@@

\usepackage{url}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HE" in caps to complete the first word.
\IEEEPARstart{T}{he} challenge presented to scientific software developers by
the drive towards Exascale computing is considerable.

\subsection{The `shallow' Program}
For this work we have used a benchmarking program that solves the
shallow-water equations on a bi-periodic plane following the
finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
Originally written by Paul Swarztrauber of the National Center for
Atmospheric Research, US.  Organic development - optimised for
previous hardware. In describing our work, we term the version of the
shallow code obtained at the beginning of this work the `original'
version.

@@

Since any real oceanographic computational model must output results,
we ensure that any \psykal version of shallow retains the Input/Output
capability of the original. This aids in limiting the optimisations
that can be performed on the \psykal version to those that should also
be applicable to full oceanographic models. Note that although we
retain the I/O functionality, all of the results presented in this work
carefully exclude the effects of I/O since it is compute performance
that interests us here.

@@

In order to maximise the flexibility (and thus potential for
optimisation) of the \psykal version of shallow, we made the kernels
as fine-grained as possible. In this case, this resulted in eight
distinct kernels, each of which operated on a single field at a single
point (since we have chosen to use point-wise kernels).

\begin{figure}
\centering
\includegraphics[width=85mm]{psykal_shallow}
\caption{A schematic of the vanilla \psykal version of the shallow code.}
\label{FIG_psykal_shallow_structure}
\end{figure}

@@

With the Gnu compiler (first and third clusters in
Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal version of
shallow only achieves ~5-6\% of the performance of the
original. Simply copying the code for each of the kernel subroutines
into the module from which they are called has the most dramatic
effect on the performance of the compiled code; it now achieves almost
80\% of the performance of the original. The next most significant
transformation is that of (manually) in-lining the kernel code itself
into the bodies of the loops from which they are called. This gets the
\psykal version to within 8\% of the performance of the original.  The
dramatic performance improvement obtained from the first
transformation demonstrates that the Gnu compiler is unable to
optimise over separate files. Therefore, the fact that the
time-stepping loop in the \psykal version still retains a call to an
I/O routine and a call to the middle-layer, both of which are in
different files, is likely to account for this remaining performance
penalty of 8\%.

@@

! Copy new fields into current fields
DO J=1,N+1
  DO I=1,M+1
    Ufld(I,J) = UNEW(I,J)
  END DO
END DO
\end{verbatim}
\caption{Example of the coding structure that required manual
  in-lining of the kernel body (as indicated by commented-out lines)
  to retrieve the performance of the original version of shallow with
  the Intel compiler.}
\label{FIG_time_smooth_code}
\end{figure}
